The Name of the Title Is Hope

### 2025-01-20 paper.tex
Add Implementation Section. Refine FlowChart.

\subsubsection{Reduce-then-Scan}
If memory bandwidth resources are sufficiently scarce relative to compute, memory-bound algorithms like scan can profit by trading additional computation for reduced memory traffic. This trade-off led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{Merrill-Grimshaw, Ha-and-Han, Dotsenko, Breitbart}: by writing only the work tile reduction to global memory in the first phase and performing a redundant scan in the third phase, $n$ global data movement can be eliminated. Like \emph{Scan-then-Propagate}, \emph{Reduce-then-Scan} employs a three-phase approach using kernel launches to resolve dependencies. However, implementations of \emph{Reduce-then-Scan} introduced new ways of processing arbitrarily-large inputs that improved over the previous recursive technique.
\begin{enumerate}
  \item \textbf{Work Tile Reduction:} Each workgroup performs a pure reduction on its work tile partition, posting the result to global memory.
  \item \textbf{Spine Scan:} The spine is gathered and scanned to produce the root.
  \item \textbf{Intra-Workgroup Scan and Propagation:} Each workgroup performs the local scan on the work tile, incorporating its portion of the root as it writes to global memory.
\end{enumerate}

@@

\subsection{Why does \emph{Chained-Scan} Rely on Forward-Progress Guarantees?}
The challenges faced by \emph{Chained-Scan} are analogous to the \emph{Dining Philosophers Problem}, where tasks compete for shared resources, potentially resulting in deadlock. Coffman et al.~\cite{10.1145/356586.356588} define four conditions necessary for deadlock:
\begin{enumerate}
  \item \textbf{Mutual Exclusion}: Tasks claim exclusive control of resources they require.
  \item \textbf{Resource Holding}: Tasks hold acquired resources while waiting.
  \item \textbf{No Preemption}: A resource cannot be removed from a task before completion.
  \item \textbf{Circular Dependency}:
  \begin{enumerate}
    \item Dependency Relationships: A dependency exists such that each task holds one or more resources that are required by proceeding tasks.
    \item Circular Ordering: A circular chain of tasks exists.
  \end{enumerate}
\end{enumerate}
Applying this schema to \emph{Chained-Scan}, workgroups correspond to "tasks" and work-tile reductions to "resources." Conditions 1-3 are easily satisfied due to the one-to-one mapping of work-tiles to workgroups. Similarly, condition 4a is met by the dependency relationships between work tiles, leaving condition 4b---circular ordering---as the key concern.

Although the serial assignment of work tiles ensures that predecessor work tiles are either scheduled or completed, there is no guarantee that the corresponding reductions will be available when a workgroup begins its lookback. This uncertainty arises because workgroups are scheduled individually \thomas{Possibly confusing. . . and perhaps not even true? Not in lockstep?}, and a predecessor workgroup might not yet have had the opportunity to compute and post its reduction. In such cases, the dependent workgroup enters a waiting state, spinning until the reduction becomes available. During this waiting period, the spinning workgroup remains scheduled, consuming computational resources. Without \emph{forward-progress guarantees}, the scheduler may schedule the spinning workgroup indefinitely over its predecessor, preventing the predecessor from completing its reduction. This satisfies condition 4b, creating a circular dependency that results in deadlock within the kernel.

@@

\subsection{Earlier Attempts at Portability}
Levien and Naur~\cite{Raph2021} made the first attempt at implementing a \emph{DecoupledLookback} approach without forward progress requirements in \emph{ScalarFallback}. By allowing a workgroup to access and operate on another tile’s data, \emph{ScalarFallback} targets Coffman et al.'s \emph{mutual exclusion} condition. Instead, each time a workgroup spins on a preceding tile, it advances a \emph{fallback} operation by incrementally consuming the blocking tile---one element at a time. Although this entails performing redundant work when the blocking tile is simply late but not deadlocking, it guarantees the termination of the algorithm because each workgroup can perform the reduction of any blocking tile, breaking the inter-workgroup dependency.

\emph{ScalarFallback} suffers from two primary issues that impact its efficiency. First, the fallback reduction is performed serially by a single thread, progressing only when the workgroup engages in a waiting spin. This effectively takes $t$ spins to complete the fallback of a deadlocking $t$-sized tile. Moreover, because only a single thread is responsible for loading elements, the cache line associated with the loaded elements is severely underutilized, further degrading performance. Second, once the fallback is complete, the workgroup does not post its result to device memory. Consequently, every workgroup deadlocked by that tile must redundantly execute the fallback, wasting computational resources. While \emph{ScalarFallback} was able to achieve speed of light on hardware supporting FPG and executed correctly on hardware without FPG, its performance on the latter was inferior to that of \emph{Reduce-Then-Scan}.

@@

\thomas{What I want to accomplish here is to go beyond saying "Here is our method. It works great." I want to show that these are the possible avenues to attack deadlocking . . . and exhaustively demonstrate that removing mutual exclusion is the best way to do it, under our given constraints.}
Like \emph{Scalar-Fallback}, we choose \emph{mutual exclusion} as our approach to break deadlocking. \emph{Dependency relationships} and \emph{no preemption} are immutable characteristics of scan and the GPU programming model, eliminating them as viable avenues for breaking deadlocking. While it may be possible to break \emph{circular ordering} by designing tile sizes such that workgroups are mapped one-to-one to multiprocessors, doing so portably is infeasible. Moreover, vendor guidance on progress guarantees among multiprocessors is also lacking, leaving the same risks of deadlocking but at the multiprocessor level. While eliminating \emph{resource holding}---allowing workgroups to discard work tiles when they encounter a deadlock---is theoretically possible, it is strictly worse than eliminating \emph{mutual exclusion}, as it necessitates $t$ additional movements (discard-read-read versus read).

To maximize the efficiency of the fallback operation, we employ three strategies. First, although a lookback is performed by a single thread or subgroup, we want all threads in the workgroup to participate. To facilitate this, we use a workgroup-wide lock to allow threads participating in the lookback to signal the entire workgroup when a fallback is necessary. Second, we explicitly limit the number of times a workgroup may spin while waiting on a tile. If the maximum spin count is exceeded, the fallback operation is initiated. Although this approach still risks performing redundant fallbacks---since a work tile that is late is indistinguishable from one that is deadlocking---exposing the maximum spin count as a tunable parameter allows us to reduce the likelihood of false-positive fallbacks. Finally, after completing a fallback, we attempt to insert the value into the spine at the position of the deadlocking tile, preventing succeeding workgroups from initiating redundant fallbacks.

The viability of \emph{Decoupled-Fallback} relies on the same cache locality properties as \emph{Decoupled-Lookback}. Regardless of FPG, due to false positives, any given hardware will have a natural fallback-incidence rate $f$. Consequently, we expect memory bandwidth traffic to increase (and throughput to decrease) by $O(fn)$ to $O((2 + f)n)$. However, just as serial ordering guarantees that memory accesses to the spine fall within the $o$-width sliding window, memory accesses to the scan input are similarly guaranteed to fall within an $ot$-width sliding window, making them also highly likely to reside within some level of the cache. As we later empirically demonstrate, this cache residency significantly reduces the cost of memory access, such that the fallback operation is almost always bound by the compute required to reduce the blocking tile. As a pure reduction incurs very lower overhead, we observe throughput approaching $O(2n)$ on most devices.

\section{Implementation}
At a high level, our algorithmn follows a five phase construction:
\begin{enumerate} 
  \item \textbf{Load Phase}: Each thread loads $p$ vectors, consuming the input in a subgroup-blocked ordering. For a subgroup size $s$ and vector size $v$, each subgroup processes a contiguous block of size $s \cdot p \cdot v$. 
  \item \textbf{Subgroup-Rake}: Each thread performs a private, serial scan across its vectors and then participates in $p$ raking subgroup scans. This allows the entire scan across the subgroup-block to be barrier-free and remain in registers. The final subgroup-block reduction is then posted to shared memory. 
  \item \textbf{Workgroup-Wide Scan}: The workgroup collectively performs a scan across the subgroup reductions, writing the results back to shared memory upon completion. 
  \item \textbf{Decoupled-Fallback}: The first subgroup in the workgroup performs a lookback operation, engaging in fallback operations if necessary. Once completed, it posts the results to shared memory for the workgroup to consume. 
  \item \textbf{Write Phase}: Each subgroup incorporates the results from the workgroup-wide scan and lookback operation to produce and write out the correct output. 
\end{enumerate}
Phases 1, 2, 3, and 5 comprise the intra-workgroup scan, following a \emph{Scan-then-Propagate} strategy. These phases are most similar to the \emph{CUDPP}~\cite{} scan kernel, with adaptations to enhance portability. Phase 4, \emph{Decoupled-Fallback}, is responsible for managing all inter-workgroup dependencies.

\subsection{Intra-Workgroup Portability Adaptations}
As our loading and writing phase are typical of scan kernels, we focus here on our portability-specific adaptations. 
\subsubsection{Subgroup-Raking}
We desire the smallest intra-workgroup spine possible. Not only does this reduce shared memory usage, but it also directly impacts the number of workgroup barriers encountered during the workgroup-wide scan, as this number is proportional to the size of the local spine. While large subgroups are relatively insensitive to this effect, small subgroups are particularly affected (reference depth here). To address this, we perform $p$ raking iterations of subgroup Kogge-Stone scans, where the highest-ranked thread passes its output into each succeeding scan. For a workgroup of size $w$, this approach reduces the number of reductions that each subgroup must post from $p$ to $1$, and it decreases the size of the spine from $\frac{pw}{s}$ to $\frac{w}{s}$.
\begin{algorithm}[htbp]
  \small
  \SetAlgoLined
  \KwIn{Shared Memory Array to Scan $x$, Workgroup Size $W$, Subgroup Size $S$}
  \KwOut{Shared Memory Inclusive Scanned Array $x$}

@@

        \If{$cond1 \textbf{ \&\& } cond2$}{
          $x[fanout\_index] \gets x[fanout\_index] + x[((fanout\_index / stride) + top\_offset) - 1]$\;
        }
      }
      $stride \gets stride * S$\;
      $top\_offset \gets top\_offset + step$\;
    }
    \textbf{barrier()}\;
  }
  \Return{$x$}\;
  \caption{Workgroup-Wide Scan}
  \label{alg:example}
\end{algorithm}
\subsubsection{Workgroup-Wide Scan}
The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$; on some hardware, subgroup sizes can vary between kernel launches. To handle this variability, we use a generalized radix-$s$ Ladner-Fischer scan network embedded with Kogge-Stone subgroup scans to perform the scan across the local spine. Because the upsweep phase of the Ladner-Fischer network follows a Brent-Kung~\cite{1675982} construction, assuming shared memory bank width equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts. To mitigate this, we employ a Merrill-Grimshaw~\cite[Section 3.3.5]{Merrill2009} style conflict avoidance strategy, which reduces bank conflicts to the radix base of the network—in our case, $s$. Counterintuitively, the resulting $s$-way conflict differs from the conflicts incurred by the Brent-Kung construction. Rather than multiple threads accessing different indices in the same memory bank, this conflict arises from a fanout originating from a single index. However, such access is interpreted by the hardware as a broadcast operation, resulting in no conflicts. Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; zero bank conflicts across all supported subgroup sizes. \footnote{Since we also require a workgroup-wide pure reduce operation, this can be achieved by omitting the fanout operation and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.}

\subsection{Decoupled-Fallback}
To reiterate, the goal of a lookback is twofold: to determine the reduction of all preceding work tiles and to communicate those values to succeeding tiles to accelerate their own lookback operations. Following the \emph{Decoupled-Lookback} technique, we model the spine as a finite state machine, where each work tile comprising the spine may exist in one of three states:
\begin{itemize}
    \item \textbf{Not Ready}: The tile has not yet been processed or written to global memory.
    \item \textbf{Ready}: The tile has completed its local reduction and posted it to global memory.
    \item \textbf{Inclusive}: The tile has completed its lookback and updated its posting to reflect the inclusive reduction.
\end{itemize}
Each lookback is, therefore, a traversal from the \emph{Ready} state to the \emph{Inclusive} state. To illustrate this process, we present a flowchart (SEE LOOKBACK FLOWCHART). Prior to the kernel launch, all tiles in the spine are initialized to the \emph{Not Ready} state. As shown in the dotted section of the flowchart, fallbacks occur when a workgroup encounters a blocking tile and the maximum spin count is exceeded.

\subsubsection{Tile State Implementation}
To maintain a coherent view of the tile's state, two bits are required, corresponding to the three possible states described earlier. In \emph{Decoupled-Lookback}, Merrill and Garland propose two strategies for communicating tile state within a 32-bit value: Use two separate 32-bit values—one for the tile state and the other for the reduction value—combined with a \emph{memory fence}, ensuring that updating the tile state also synchronizes the reduction value. Use a single 64-bit value by bit-packing the tile state and reduction value, with updates managed via 64-bit atomics.

In our case, neither are viable options. Due to a lack of support across backend APIs, WebGPU does not furnish developers with a memory model, and currently, 64-bit atomics are not supported. Moreover, 64-bit atomics are not supported across most mobile devices, and thus we do not consider their future support an appropriate option for our implementation. Instead, we use two 32-bit values and split the reduction into upper and lower 16-bit segments, reserving the two most significant bits of each segment to encode the tile state.
\newline
SPLIT BITS MINI DIAGRAM here
\newline

\begin{enumerate}
  \item Idea: Split the reductions into multiple values; use separate values for local/inclusive instead of using flags to differentiate them.
  \item Pro: Full 32 or theoretically 64 bits.
  \item Pro: No memory model or 64 bit atomics required!
  \item Pro: Can safely fallback insert without risking clobbering.
  \item Con: 2x Memory usage
  \item Con: Places subgroup functions adjacent to atomics, which may lead to \textbf{\textit{interesting}} behavior.
\end{enumerate}

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{graphics/FlowChart.pdf}
  \caption{TODO update for split}
\end{figure}

\subsubsection{Full Algorithm}

\begin{algorithm}[htbp]
  \tiny
  \SetAlgoLined
  \DontPrintSemicolon

  \KwIn{%
    \emph{Partition ID} $part\_id$, 
    \emph{Number of lanes} $lane\_count$, 
    \emph{Workgroup-shared lock} $wg\_lock$, 
    \emph{Atomic reduction array} \texttt{reduction}[], 
    \emph{Local partial reductions} $wg\_reduce[]$, 
    \emph{Constants:} $MAX\_SPIN\_COUNT, FLAG\_MASK, FLAG\_NOT\_READY, FLAG\_INCLUSIVE, FLAG\_READY, ALL\_READY, \dots$
  }
  \KwOut{Possible updates to $prev\_red$ or partial progress before fallback.}

  \If(\tcp*[h]{Only needed if not the first partition}){$part\_id \neq 0$}{
    $\,prev\_red \gets 0$\;
    $\,lookback\_id \gets part\_id - 1$\;
    $\,lock \gets \texttt{workgroupUniformLoad}(\&wg\_lock)$\;

    \While(\tcp*[h]{Spin until we can proceed or must fallback}){$lock = \textnormal{LOCKED}$}{
      \If(\tcp*[h]{Only threads within the subgroup size do checks}){$threadid.x < lane\_count$}{
        $\,spin\_count \gets 0$\;
        \While{$spin\_count < MAX\_SPIN\_COUNT$}{
          $\,flag\_payload \gets \texttt{select}\bigl(
             0,\;
             \texttt{atomicLoad}(\&\texttt{reduction}[\,lookback\_id\,][\,threadid.x\,]),\;
             threadid.x < \textnormal{SPLIT\_MEMBERS}
           \bigr)$\;

          \If{$\texttt{subgroupBallot}\bigl((\,flag\_payload \,\&\, FLAG\_MASK) > FLAG\_NOT\_READY\bigr).x = ALL\_READY$}{
            $\,incl\_bal \gets \texttt{subgroupBallot}\bigl(
               (\,flag\_payload \,\&\, FLAG\_MASK) = FLAG\_INCLUSIVE
             \bigr).x$\;

>>>>

\subsubsection{Reduce-then-Scan}
If memory bandwidth resources are sufficiently scarce relative to compute, memory-bound algorithms like scan can profit by trading additional computation for reduced memory traffic. This trade-off led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{Merrill-Grimshaw, Ha-and-Han, Dotsenko, Breitbart}: by writing only the work tile reduction to global memory in the first phase and performing a redundant scan in the third phase, $n$ global data movement can be eliminated. Like \emph{Scan-then-Propagate}, \emph{Reduce-then-Scan} employs a three-phase approach using kernel launches to resolve dependencies.
\begin{enumerate}
  \item \textbf{Work Tile Reduction:} Each workgroup performs a pure reduction on its work tile partition, posting the result to global memory.
  \item \textbf{Spine Scan:} The spine is gathered and scanned to produce the root.
  \item \textbf{Intra-Workgroup Scan and Propagation:} Each workgroup performs the local scan on the work tile, incorporating its portion of the root as it writes to global memory.
\end{enumerate}

@@

\subsection{Why does \emph{Chained-Scan} Rely on Forward-Progress Guarantees?}
The challenges faced by \emph{Chained-Scan} are analogous to the \emph{Dining Philosophers Problem}, where tasks compete for shared resources, potentially resulting in deadlock. Coffman et al.~\cite{10.1145/356586.356588} define four conditions necessary for deadlock:
\begin{enumerate}
  \item \textbf{Mutual Exclusion}: Tasks claim exclusive control of resources they require.
  \item \textbf{Resource Holding}: Tasks hold acquired resources while waiting.
  \item \textbf{No Preemption}: A resource cannot be removed from a task before completion.
  \item \textbf{Circular Dependency}:
  \begin{enumerate}
    \item Dependency Relationships: A dependency exists such that each task holds one or more resources that are required by proceeding tasks.
    \item Circular Ordering: A circular chain of tasks exists.
  \end{enumerate}
\end{enumerate}
Applying this schema to \emph{Chained-Scan}, workgroups correspond to "tasks" and work-tile reductions to "resources." Conditions 1–3 are satisfied due to the one-to-one mapping of work tiles to workgroups and the constraints of the GPU programming model. Similarly, condition 4a is met by the dependency relationships between work tiles, leaving condition 4b---circular ordering---as the key concern.

Although the serial assignment of work tiles ensures that predecessor work tiles are either scheduled or completed, there is no guarantee that the corresponding reductions will be available before a workgroup begins its lookback. This uncertainty arises because workgroups are scheduled individually, and a predecessor workgroup might not yet have had the opportunity to compute and post its reduction. In such cases, the dependent workgroup enters a waiting state, spinning until the reduction becomes available. During this waiting period, the spinning workgroup remains scheduled, consuming computational resources. Without \emph{forward-progress guarantees}, the scheduler may indefinitely prioritize the spinning workgroup over its predecessor, preventing the predecessor from completing its reduction. This creates a circular dependency, satisfying condition 4b and resulting in deadlock within the kernel.

@@

\subsection{Earlier Attempts at Portability}
Levien and Naur~\cite{Raph2021} made the first attempt at implementing a \emph{DecoupledLookback} approach without forward progress requirements in \emph{ScalarFallback}. By allowing a workgroup to access and operate on another tile’s data, \emph{ScalarFallback} targets Coffman et al.'s \emph{mutual exclusion} condition. Instead, each time a workgroup spins on a preceding tile, it advances a \emph{fallback} operation by incrementally consuming the blocking tile---one element at a time. Although this entails performing redundant work when the blocking tile is simply late but not blocking, it guarantees the termination of the algorithm because each workgroup can perform the reduction of any blocking tile, breaking the inter-workgroup dependency.

\emph{ScalarFallback} suffers from two primary issues that impact its efficiency. First, the fallback reduction is performed serially by a single thread, progressing only when the workgroup engages in a waiting spin. This effectively takes $t$ spins to complete the fallback of a blocking $t$-sized tile. Moreover, because only a single thread is responsible for loading elements, the cache line associated with the loaded elements is severely underutilized, further degrading performance. Second, once the fallback is complete, the workgroup does not post its result to device memory. Consequently, every workgroup deadlocked by that tile must redundantly execute the fallback, wasting computational resources. While \emph{ScalarFallback} was able to achieve speed of light on hardware supporting FPG and executed correctly on hardware without FPG, its performance on the latter was inferior to that of \emph{Reduce-Then-Scan}.

@@

\thomas{What I want to accomplish here is to go beyond saying "Here is our method. It works great." I want to show that these are the possible avenues to attack deadlocking . . . and exhaustively demonstrate that removing mutual exclusion is the best way to do it, under our given constraints.}
Like \emph{Scalar-Fallback}, we choose \emph{mutual exclusion} as our approach to break deadlocking. \emph{Dependency relationships} and \emph{no preemption} are immutable characteristics of scan and the GPU programming model, eliminating them as viable avenues for breaking deadlocking. While it may be possible to break \emph{circular ordering} by designing tile sizes such that workgroups are mapped one-to-one to multiprocessors, doing so portably is infeasible. Moreover, vendor guidance on progress guarantees among multiprocessors is also lacking, leaving the same risks of deadlocking but at the multiprocessor level. While eliminating \emph{resource holding}---allowing workgroups to discard work tiles when they encounter a deadlock---is theoretically possible, it is strictly worse than eliminating \emph{mutual exclusion}, as it necessitates $t$ additional movements (discard-read-read versus read).
\begin{algorithm}[htb]
  \small
  \SetAlgoLined
  \KwIn{%
    \texttt{in}[], \texttt{out}[], 
    \texttt{tileBump}, 
    \texttt{spine}[], 
    \texttt{wg\_fallback}[], 
    \texttt{wg\_lock}, 
    \texttt{wg\_partials}[], 
    \texttt{wg\_prev\_reduction},
    \texttt{wg\_broadcast}, 
  }
  \KwOut{\texttt{out}[] fully inclusive scan of \texttt{in}[]}

  \If{$\textnormal{threadid.x} = 0$}{
    \texttt{wg\_broadcast} $\gets$ \texttt{atomicAdd}(\&\texttt{tileBump}, 1)\;
    \texttt{wg\_lock} $\gets$ \textnormal{LOCKED}\;
  }
  \textbf{barrier()}\;

  \texttt{part\_id} $\gets$ \texttt{wg\_broadcast}\;

  \textit{local\_vectors}: $\textnormal{array}\langle \textnormal{vec4}\langle u32\rangle, p\rangle$\;

  \textbf{Load}(\texttt{in}, \texttt{part\_id}, \textit{local\_vectors})\;
  \textbf{SubgroupRake}(\textit{local\_vectors})\;

  \If{$\text{\normalfont highestRankLane}(\textnormal{threadid.x})$}{
    \texttt{wg\_partials}[\textnormal{subgroup\_id}] $\gets$ \textit{local\_vectors}[\textnormal{final\_partial}]\;
  }
  \textbf{barrier()}\;

  \textbf{WorkgroupWideScan}(\texttt{wg\_partials})\;
  \textbf{barrier()}\;

  \tcp{Decoupled Fallback}
  \While{\texttt{wg\_lock} = \textnormal{LOCKED}}{
    \textbf{Lookback}(\texttt{spine}, \texttt{wg\_prev\_reduction}, \texttt{wg\_lock})\;
    \texttt{lock} $\gets$ \textbf{\texttt{workgroupUniformLoad}}(\&\texttt{wg\_lock})\;

    \If{$\texttt{lock} = \textnormal{LOCKED}$}{
      \textbf{FallbackReduce}(\texttt{in}, \texttt{wg\_fallback})\;
      \textbf{FallbackInsertionAttempt}(\texttt{spine}, \texttt{wg\_prev\_reduction}, \texttt{wg\_lock})\;
      \texttt{lock} $\gets$ \textbf{\texttt{workgroupUniformLoad}}(\&\texttt{wg\_lock})\;
    }
  }

  \textbf{Write}(\texttt{out}, \textit{local\_vectors}, \texttt{wg\_partials}, \texttt{wg\_prev\_reduction}, \texttt{part\_id})\;

  \caption{High-level Decoupled Fallback}
\end{algorithm}
To maximize the efficiency of the fallback operation, we employ three strategies. First, although a lookback is performed by a single thread or subgroup, we want all threads in the workgroup to participate. To facilitate this, we use a workgroup-wide lock to allow threads participating in the lookback to signal the entire workgroup when a fallback is necessary. Second, we explicitly limit the number of times a workgroup may spin while waiting on a tile. If the maximum spin count is exceeded, the fallback operation is initiated. Although this approach still risks performing redundant fallbacks---since a work tile that is late is indistinguishable from one that is blocking---exposing the maximum spin count as a tunable parameter allows us to reduce the likelihood of false-positive fallbacks. Finally, after completing a fallback, we attempt to insert the value into the spine at the position of the blocking tile, preventing succeeding workgroups from initiating redundant fallbacks.

The viability of \emph{Decoupled-Fallback} relies on the same cache locality properties as \emph{Decoupled-Lookback}. Regardless of FPG, due to false positives, any given hardware will have a natural fallback-incidence rate $f$. Consequently, we expect memory bandwidth traffic to increase (and throughput to decrease) by $O(fn)$ to $O((2 + f)n)$. However, just as serial ordering guarantees that memory accesses to the spine fall within an $o$-width sliding window, memory accesses to the scan input are similarly guaranteed to fall within an $ot$-width sliding window, making them also highly likely to reside within some level of the cache. As we later empirically demonstrate, this cache residency significantly reduces the cost of memory access, such that the fallback operation is almost always bound by the compute required to reduce the blocking tile. As a pure reduction incurs very lower overhead, we observe throughput approaching $O(2n)$ on most devices.

\section{Implementation}
At a high level, our algorithmn follows a five phase construction:
\begin{enumerate}
  \item[(0)] \textbf {Initialization}: The first thread in the workgroup atomically acquires its work-tile by atomically incrementing a value in global memory.
  \item \textbf{Load Phase}: Each thread loads $p$ vectors, consuming the input in a subgroup-blocked ordering. For a subgroup size $s$ and vector size $v$, each subgroup processes a contiguous block of size $s \cdot p \cdot v$. 
  \item \textbf{Subgroup-Rake}: Each thread performs a private, serial scan across its vectors and then participates in $p$ raking subgroup scans. This allows the entire scan across the subgroup-block to be barrier-free and remain in registers. The final subgroup-block reduction is then posted to shared memory. 
  \item \textbf{Workgroup-Wide Scan}: The workgroup collectively performs a scan across the subgroup partial reductions, writing the results back to shared memory upon completion. 
  \item \textbf{Decoupled-Fallback}: The first subgroup in the workgroup performs a lookback operation, engaging in fallback operations if necessary. Once completed, it posts the results to shared memory for the workgroup to consume. 
  \item \textbf{Write Phase}: Each subgroup incorporates the results from the workgroup-wide scan and lookback operation to produce and write out the correct output. 
\end{enumerate}
Phase 0 ensures that the work tiles are consumed in serial order. Phases 1, 2, 3, and 5 comprise the intra-workgroup scan, following a \emph{Scan-then-Propagate} strategy. Phase 4, \emph{Decoupled-Fallback}, is responsible for managing all inter-workgroup dependencies.

\subsection{Intra-Workgroup Portability Adaptations}
As our loading and writing phase are typical of scan kernels, we focus here on our portability-specific adaptations. 
\subsubsection{Subgroup-Raking}
We desire the smallest intra-workgroup spine possible. Not only does this reduce shared memory usage, but it also directly impacts the number of workgroup barriers encountered during the workgroup-wide scan, as this number is proportional to the size of the local spine. While large subgroups are relatively insensitive to this effect, small subgroups are particularly affected (reference depth here). To address this, we perform $p$ raking iterations of subgroup Kogge-Stone scans, where the highest-ranked thread passes its output into each succeeding scan. For a workgroup of size $w$, this approach reduces the number of partial reductions that each subgroup must post from $p$ to $1$, and it decreases the size of the spine from $\frac{pw}{s}$ to $\frac{w}{s}$.
\begin{algorithm}[htbp]
  \small
  \SetAlgoLined
  \KwIn{Shared Memory Array to Scan $x$, Workgroup Size $W$, Subgroup Size $S$}
  \KwOut{Shared Memory Inclusive Scanned Array $x$}

@@

        \If{$cond1 \textbf{ \&\& } cond2$}{
          $x[fanout\_index] \gets x[fanout\_index] + x[((fanout\_index / stride) + top\_offset) - 1]$\;
        }
      }
      $stride \gets stride * S$\;
      $top\_offset \gets top\_offset + step$\;
    }
    \textbf{barrier()}\;
  }
  \Return{$x$}\;
  \caption{Workgroup-Wide Scan}
  \label{alg:example}
\end{algorithm}
\subsubsection{Workgroup-Wide Scan}
The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$; on some hardware, subgroup sizes can vary between kernel launches. To handle this variability, we use a generalized radix-$s$ Ladner-Fischer scan network embedded with Kogge-Stone subgroup scans to perform the scan across the local spine. Because the upsweep phase of the Ladner-Fischer network follows a Brent-Kung~\cite{1675982} construction, assuming shared memory bank width equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts. To mitigate this, we employ a Merrill-Grimshaw~\cite[Section 3.3.5]{Merrill2009} style conflict avoidance strategy, which reduces bank conflicts to the radix base of the network—in our case, $s$. Counterintuitively, the resulting $s$-way conflict differs from the conflicts incurred by the Brent-Kung construction. Rather than multiple threads accessing different indices in the same memory bank, this conflict arises from a fanout originating from a single index. However, such access is interpreted by the hardware as a broadcast operation, resulting in no conflicts. Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase is negligible due to our overall thriftiness, including the deliberate retention of vectors in registers and the efficiency of subgroup raking. Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; zero bank conflicts across all supported subgroup sizes. \footnote{Since we also require a workgroup-wide pure reduce operation, this can be achieved by omitting the fanout operation and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.}

\subsection{Decoupled-Fallback}
To reiterate, the goal of a lookback is twofold: to determine the reduction of all preceding work tiles and to communicate those values to succeeding tiles to accelerate their own lookback operations. Following the \emph{Decoupled-Lookback} technique, we model the spine as a finite state machine, where each work tile comprising the spine may exist in one of three states:
\begin{itemize}
    \item \textbf{Not Ready}: The tile has not yet been processed or written to global memory.
    \item \textbf{Ready}: The tile has completed its local reduction and posted it to global memory.
    \item \textbf{Inclusive}: The tile has completed its lookback and updated its posting to reflect the inclusive reduction.
\end{itemize}
Each lookback is, therefore, a traversal from the \emph{Ready} state to the \emph{Inclusive} state. To illustrate this process, we present a flowchart (SEE LOOKBACK FLOWCHART).

\subsubsection{Decoupled-Fallback}
\thomas{Given that "lock" has an extensive history of use in computer science and this is not that kind of lock, perhaps a better choice of phrase would be better. Updates to terminology incoming...}
Prior to kernel launch, all tiles in the spine are initialized to the \emph{Not Ready} state. During partition index acquisition, we also initialize a shared memory \emph{control flag} to facilitate synchronization within the workgroup during the \emph{Decoupled-Fallback} phase. This flag governs both the initiation of fallback operations and the entry and exit of the entire phase, which begins immediately after the workgroup-wide scan over partial reductions is completed.
\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{graphics/FlowChart.pdf}
  \caption{Decoupled-Fallback}
\end{figure}
\begin{enumerate}
  \item[(0)] \textbf{Posting Local Reduction:} Each workgroup posts its local reduction---the result of the workgroup-wide scan over partials---to global memory and updates its tile state to \emph{Ready}. If it is the first work tile, its local reduction \emph{is} its inclusive reduction. In this case, the workgroup updates its tile state directly to \emph{Inclusive} and bypasses the lookback entirely.

  \item \textbf{Initialization:}
  \begin{itemize}
      \item The control flag, initialized to \emph{Locked}, signals that the workgroup should enter the lookback phase.
      \item Each workgroup initializes a \emph{previous reduction} variable with the identity element of the monoid to hold the accumulated reductions of preceding tiles.
      \item The workgroup prepares to traverse the spine, starting with its immediate predecessor tile.
  \end{itemize}

  \item \textbf{Lookback:}
  \begin{itemize}
      \item The workgroup queries the state of the current tile. 
      \item If the tile is in the \emph{Ready} state, the workgroup accumulates the value and moves to the preceding tile, repeating this step.
      \item If the tile is in the \emph{Inclusive} state, the workgroup accumulates the value and proceeds directly to step 5.
      \item If the tile is in the \emph{Not Ready} state, the workgroup spins for a limited number of iterations, defined by the \emph{Maximum Spin Count}. If the spin count exceeds the \emph{Maximum Spin Count}, the workgroup signals a fallback operation using the control flag and proceeds to the next step.
  \end{itemize}

  \item \textbf{Fallback:}
  \begin{itemize}
      \item The workgroup computes the reduction of the blocking tile by processing the corresponding elements directly.
      \item Once the fallback reduction is complete, the workgroup \emph{attempts} to post the reduction to global memory and update the tile's state to \emph{Ready}.
      \item As this update attempt is performed atomically, the workgroup reads back the blocking tile's immediate state:
      \begin{itemize}
          \item If the state is \emph{Inclusive}, the fallback value is discarded, and the workgroup accumulates the value before proceeding to step 5.
          \item If the blocking tile is the first tile in the spine, the workgroup has reached the start of the spine and proceeds to step 5.
          \item Otherwise, the workgroup resumes the lookback, starting with the preceding tile.
      \end{itemize}
  \end{itemize}

  \item \textbf{Exit:}
  \begin{itemize}
      \item Once the lookback or fallback is complete, the workgroup posts its inclusive reduction to global memory, transitioning its tile state to \emph{Inclusive}.
      \item The \emph{previous reduction} is also posted to shared memory for the workgroup to consume and propagate as necessary.
      \item The control flag is updated to \emph{Unlocked} to signal the exit of the \emph{Decoupled-Fallback} phase, allowing subsequent operations to proceed.
  \end{itemize}
\end{enumerate}

\subsubsection{Tile State Implementation}
To maintain a coherent view of the tile's state, two bits are required. In \emph{Decoupled-Lookback}, Merrill and Garland propose two strategies for communicating tile state and 32-bit reduction value: Two separate 32-bit values---one for the tile state and the other for the reduction value---with updates using a \emph{memory fence}. One single 64-bit value, bit-packing the tile state and reduction value, with updates via 64-bit atomics.
In our case, neither are viable options. Due to a lack of support across backend APIs, WebGPU does not furnish developers with a memory model, and currently, 64-bit atomics are not supported. Moreover, 64-bit atomics are not supported across most mobile devices, and thus we do not consider their future support an appropriate option for our implementation. Instead, we use two 32-bit values and split the reduction into upper and lower 16-bit segments, reserving the two most significant bits of each 32-bit value to encode the tile state.
\newline
SPLIT BITS MINI DIAGRAM here
\newline
To accommodate split values, we use a single subgroup for the lookback, leveraging subgroup \emph{ballot} to verify that the split value state is consistent across threads, and subgroup \emph{shuffle} to join and split reductions as needed. Since this method requires one thread per 16 bits, and the minimum subgroup size in the WebGPU specification is 4, it supports both 32-bit and 64-bit values. Notably, our approach extends support to 64-bit scans, which would not be possible even with 64-bit atomics.\footnote{Although native 64-bit values are not currently supported by the WebGPU specification, their future inclusion is anticipated.}
